Wormtable: A data structure for genome scale data

### 2013-07-23 docs/paper/paper.tex

\title[Wormtable]{Wormtable: A data structure for genome scale data}
\author[Kelleher \textit{et~al}]{Jerome Kelleher, Robert W. Ness 
and Daniel L. Halligan}
\address{
University of Edinburgh,
King's Buildings,
West Mains Road,
EH9 3JT,
UK
}

@@

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of relational databases,
wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is an open source embedded database toolkit that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. Berkeley DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads and complexity of a relational database system.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF\footnote{%
\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
a Python VCF parser.
In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 130 minutes;
the wormtable version required
55 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 77 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

@@

\section*{Acknowledgement}
\paragraph{Funding\textcolon}  
JK is supported by EPSRC grant EP/I013091/1; DLH is supported by XXXXX; 
and RWN is supported by XXXX.

>>>>

\title[Wormtable]{Wormtable: A data structure for genome scale data}
\author[Kelleher \textit{et~al}]{Jerome Kelleher, Rob W. Ness 
and Daniel L. Halligan}
\address{
University of Edinburgh,
King's Buildings,
West Mains Road,
EH9 3JT,
UK
}

@@

% why do we use Berkeley DB?
To take advantage of the data processing capabilities of database
technologies without the overheads and disadvantages of relational databases,
wormtable uses Berkeley DB~\citep{obs99} for row storage and indexing.
Berkeley DB is a free and open-source embedded database toolkit that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}. Berkeley DB is a highly modular library, providing the 
means to define a very lightweight read-only database, without the 
overheads and complexity of a relational database system.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF\footnote{%
\href{http://pyvcf.rtfd.org/}{http://pyvcf.rtfd.org/}}, 
a Python VCF parser.
In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes;
the wormtable version required
55 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 77 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that this simple task represents 
something of a best case scenario for the Unix tools pipeline.

@@

\section*{Acknowledgement}
We would like to thank the Ashworth Codemonkeys for valuable feedback on earlier
versions of wormtable.
\paragraph{Funding\textcolon}  
JK is supported by EPSRC grant EP/I013091/1.
